Scoring Patients

### 2018-05-02 AllSamplesModel.Rmd
Moved shiny app to own repo

hazards = apply( exp(t(dysplasia.df[, rownames(coefs[[select.alpha]])]) *  coefs[[select.alpha]][,1]), 1, function(x) {
  sd(x)/mean(x)
})
ch = as.data.frame(hazards)
ch$coef = coefs[[select.alpha]][,1]
ch$stability = featureStability
ch$label = rownames(ch)
ch = arrange(ch, -hazards)

@@

## Top Features
```{r, echo=T}
m = data.frame(matrix(nrow=nrow(chh), ncol=0))

@@

@@

One way of looking at the p53 prediction (e.g. Adam Bass) is to see how predictive the p-arm loss in chr17 is.
```{r, echo=F, warning=F, message=F, fig.height=5, fig.width=10}
ccgenes = read.table('~/Data/CosmicCensusGenes.tsv', sep='\t', header=T, stringsAsFactors=F)
ccgenes = ccgenes[-grep('^X|Y', ccgenes$Genome.Location),]
#ccgenes = ccgenes[-which(ccgenes$Mutation.Types == 'T'),] # drop the translocations
#ccgenes = subset(ccgenes, !grepl('^X|Y', Genome.Location))
#loc = unlist(strsplit(ccgenes$Genome.Location, ':|-'))
loc = do.call(rbind.data.frame, strsplit(ccgenes$Genome.Location, ':|-'))
colnames(loc) = c('chr','start','end')
loc[] = lapply(loc[], function(x) as.numeric(as.character(x)))

@@

cols = c('Hospital.Research.ID','Patient','Status','PID', 'Samplename','Endoscopy.Year','Pathology','Prediction','OR')
dspred = dspred %>% dplyr::mutate( 'Adj. RR'=RR+offsetMean )

dspred = dspred %>% dplyr::mutate( 'Adj. Prob'=1/(1+exp(-RR+abs(offsetMean)) ))

ggplot(dspred, aes(Type, `Adj. Prob`)) + geom_jitter()

ggplot(dspred, aes(Type, `Adj. RR`)) + geom_jitter() + geom_boxplot(outlier.color = NA)

head(dspred)
m = melt(dspred, measure.vars=c('RR','Adj. RR'))
colors = myPal[seq(2, length(myPal), round(length(myPal)/4))]
pr = ggplot(m, aes(variable, value, fill=Type)) + facet_grid(~Type, scales='fixed') + 
  geom_jitter(width=0.2, aes(color=value)) + 
  scale_color_gradientn(colors = myPal,  name='') + 
  geom_boxplot(outlier.colour = NA) + 
  scale_fill_manual(values = colors,  name='') + 
  labs(x='', y='Relaive Risk', title='Comparing Relative Risk') + 
  theme(legend.position = 'none', axis.text.x = element_text(angle=45,hjust=1)) + plot.theme 
pr
ggsave('plots/downsampled_adjustedRR_supplement.png', plot=pr, width=7, height=5,dpi=300, scale=1)

>>>>

hazards = apply( exp(t(dysplasia.df[,rownames(coefs[[select.alpha]])]) *  coefs[[select.alpha]][,1]), 1, function(x) {
  sd(x)/mean(x)
  #mad(x) # The CV(RR) is pretty sensitive to extremes I think. But I'm not sure if MAD(RR would be better)
})

ch = as.data.frame(hazards)
ch$coef = coefs[[select.alpha]][,1]
ch$stability = featureStability
ch$label = rownames(ch)
ch = arrange(ch, -hazards)

@@

# Top Features
```{r, echo=F, message=F, warning=F}
m = data.frame(matrix(nrow=nrow(chh), ncol=0))

@@

write.table(chh, sep='\t', quote=F, row.names=F, file='features.txt')

@@

Clearly need to interpret those "hazards" carefully.  Some of the really high ones are driven by some of the extreme samples.

```{r, fig.width=10, fig.height=10}

gains = subset(chh, coef > 0)

y = cbind.data.frame(labels, dysplasia.df[,gains$label])
y$labels = factor(y$labels)
y$labels = plyr::revalue(y$labels, c('0'='NP','1'='P'))

m = melt(y, id.vars='labels')
colnames(m) = c('status','label','value')
m = base::merge(m, gains, by='label')
m$chr = factor(m$chr, levels=chr.lengths$chrom, ordered = T)

p = ggplot(m) + facet_wrap(chr~label, scales='free_y') +
  geom_jitter(aes(status, value, group=status, color=status), width=0.01) + 
  geom_violin(aes(status, value, group=status, fill=status, color=status), alpha=0.5) + 
  geom_hline(yintercept=c(1,-1), color='grey39', linetype='dotted') +
  geom_label(data=subset(gains, hazards >= 1), aes('P',0, label=round(hazards,2)), color='red') + 
  geom_label(data=subset(gains, hazards < 1), aes('P',0, label=round(hazards,2)), color='grey') + 
  labs(y='',x='',title='Gains') +
  plot.theme + theme(legend.position = 'none', strip.text=element_text(size=9)) 
p
ggsave('plots/gains.png', plot=p, width=10, height=10, unit='in', dpi=300)

losses = subset(chh, coef < 0)

y = cbind.data.frame(labels, dysplasia.df[,losses$label])
y$labels = factor(y$labels)
y$labels = plyr::revalue(y$labels, c('0'='NP','1'='P'))

m = melt(y, id.vars='labels')
colnames(m) = c('status','label','value')
m = base::merge(m, losses, by='label')
m$chr = factor(m$chr, levels=chr.lengths$chrom, ordered = T)

p = ggplot(m) + facet_wrap(chr~label, scales='free_y') +
  geom_jitter(aes(status, value, group=status, color=status), width=0.01) + 
  geom_violin(aes(status, value, group=status, fill=status, color=status), alpha=0.5) + 
  geom_hline(yintercept=c(1,-1), color='grey39', linetype='dotted') +
  geom_label(data=subset(losses, hazards >= 1), aes('P',3, label=round(hazards,2)), color='red') + 
  geom_label(data=subset(losses, hazards < 1), aes('P',3, label=round(hazards,2)), color='grey') + 
  labs(y='',x='',title='Losses') +
  plot.theme + theme(legend.position = 'none', strip.text=element_text(size=9)) 
p
ggsave('plots/losses.png', plot=p, width=10, height=10, unit='in', dpi=300)

One way of looking at the p53 prediction (e.g. Adam Bass) is to see how predictive the p-arm loss in chr17 is.

```{r, echo=F, warning=F, message=F, fig.height=5, fig.width=10}
ccgenes = read.table('~/Data/CosmicCensusGenes.tsv', sep='\t', header=T, stringsAsFactors=F)
ccgenes = ccgenes[-grep('^X|Y', ccgenes$Genome.Location),]
#ccgenes = ccgenes[-which(ccgenes$Mutation.Types == 'T'),] # drop the translocations
#ccgenes = subset(ccgenes, !grepl('^X|Y', Genome.Location))
#loc = unlist(strsplit(ccgenes$Genome.Location, ':|-'))
loc = do.call(rbind.data.frame, strsplit(ccgenes$Genome.Location, ':|-'))
colnames(loc) = c('chr','start','end')
loc[] = lapply(loc[], function(x) as.numeric(as.character(x)))

@@

cols = c('Hospital.Research.ID','Patient','Status','PID', 'Samplename','Endoscopy.Year','Pathology','Prediction','OR')
dspred$Illumina.ID = rownames(dspred)
dspred = dspred %>% dplyr::mutate( 'Adj. RR'=RR+offsetMean )
dspred = dspred %>% dplyr::mutate( 'Adj. Prob'=1/(1+exp(-RR+abs(offsetMean)) ))

x = 'LP6005691-DNA_A02'
subset(dspred, Illumina.ID == x)

m = melt(dspred, measure.vars=c('RR','Adj. RR'))
colors = myPal[seq(2, length(myPal), round(length(myPal)/4))]
pr = ggplot(m, aes(variable, value, fill=Type)) + facet_grid(~Type, scales='fixed') + 
  geom_jitter(width=0.2, aes(color=value)) + 
  scale_color_gradientn(colors = myPal,  name='') + 
  geom_boxplot(outlier.colour = NA) + 
  scale_fill_manual(values = colors,  name='') + 
  labs(x='', y='Relaive Risk', title='Comparing Relative Risk') + 
  theme(legend.position = 'none', axis.text.x = element_text(angle=45,hjust=1)) + plot.theme 
pr
ggsave('plots/downsampled_adjustedRR_supplement.png', plot=pr, width=7, height=5,dpi=300, scale=1)

### 2018-05-02 Patient Report.Rmd
Moved shiny app to own repo

@@

m = (melt(df, id.vars = 'status', measure.vars = 'seg.len'))
m2 = m %>% group_by(status) %>% summarise( median=median(value), n=length(status), value=max(value) )
ggplot(m, aes(status, value, group=status, fill=status)) + geom_jitter(aes(color=status), width=.2) + geom_violin(alpha=0.5) +
  geom_label(data=m2, aes(status, value, label=paste("median=",scientific(median,3),'\nn=',n, sep='')), fill='white') + 
  theme(legend.position = 'none') + labs(y='Variable segment length', x='', title=paste('Length of variable regions (all samples n=',length(patient.info$Samplename),')',sep=''), subtitle=paste('Wilcox p-value', signif(lenWT$p.value, digits=3) ))

```{r, echo=F, warning=F, message=F}
df = rbind(cbind.data.frame('n'=sapply(var.seg.lengths[prog.pt], nrow), 'status'='Progressor'), cbind.data.frame('n'=sapply(var.seg.lengths[np.pt], nrow), 'status'='Non-Progressor'))
nWT = t.test(subset(df, status == 'Progressor')$n, subset(df, status == 'Non-Progressor')$n)

ggplot(df, aes(status, n, group=status)) + geom_jitter(aes(color=status), width=0.2) +
  geom_violin(aes(fill=status, group=status), alpha=0.5) +
  geom_label(data=(df %>% group_by(status) %>% summarise(y=max(n), med=median(n))), 
             aes(status, y, label=paste('median\n',sprintf("%.1f", med, sep='')))) +
  labs(y='Number of segments', title="Number of segments per patient", subtitle=paste('t-test p-value', signif(nWT$p.value, digits=3)) ) +
  theme(legend.position = 'none')

nohgd.var.seg.lengths = lapply(patient.data, function(pt) {
  samples = subset(pt$info, Pathology %in% c('BE','ID','LGD'))$Samplename
  cols = c('chrom','start.pos','end.pos', grep(paste(samples, collapse='|'), colnames(pt$variable.regions), value=T))
  vr = pt$variable.regions[,cols]

  cbind('chr'=vr$chrom, 'start'=vr$start.pos, 'end'=vr$end.pos, 'seg.len'=vr$end.pos - vr$start.pos)
})

@@

```{r echo=F, message=F, warning=F, fig.height=8, fig.width=20}
n.var.regions = sapply(var.seg.lengths, nrow)

@@

ggplot(melted, aes(Status, value, fill=Status)) + geom_jitter(aes(color=Status), width=0.2) + geom_boxplot(outlier.colour = NA) + 
  geom_label(data=subset(melted, Status == 'P'), aes(x=Status, y=max(value)-sd(value), label=scientific(median(value),3) ), color='white', show.legend = F ) +
  geom_label(data=subset(melted, Status == 'NP'), aes(x=Status, y=max(value)-sd(value), label=scientific(median(value),3) ), color='white' , show.legend = F) +
  labs(title="Variable region length",subtitle=paste("p-value", signif(chr.ratio.tt$p.value, digits=3)), y='Segment Length/Chr length')

ggplot(melted, aes(Status, value, fill=Status)) + geom_jitter(aes(color=Status), width=0.2) + geom_violin() + 
  geom_label(data=subset(melted, Status == 'P'), aes(x=Status, y=max(value)-sd(value), label=scientific(median(value),3) ), color='white' , show.legend = F) +
  geom_label(data=subset(melted, Status == 'NP'), aes(x=Status, y=max(value)-sd(value), label=scientific(median(value),3) ), color='white' , show.legend = F) +
  labs(title="Variable region length",subtitle=paste("p-value", signif(chr.ratio.tt$p.value, digits=3)), y='Segment Length/Chr length')

>>>>

```{r echo=F, message=F, warning=F, fig.height=5, fig.width=5}
df = rbind(cbind(do.call(rbind.data.frame, var.seg.lengths[prog.pt]), 'status'='Progressor'), cbind(do.call(rbind.data.frame, var.seg.lengths[np.pt]), 'status'='Non-Progressor'))
lenWT = wilcox.test(subset(df, status == 'Progressor')$seg.len, subset(df, status == 'Non-Progressor')$seg.len)

m = (melt(df, id.vars = 'status', measure.vars = 'seg.len'))
m2 = m %>% group_by(status) %>% dplyr::summarise( median=median(value), n=length(status), value=max(value) )
ggplot(m, aes(status, value, group=status, fill=status)) + geom_jitter(aes(color=status), width=.2) + geom_violin(alpha=0.5) +
  geom_label(data=m2, aes(status, value, label=paste("median=",scientific(median,3),'\nn=',n, sep='')), fill='white') + 
  theme(legend.position = 'none') + labs(y='Variable segment length', x='', title=paste('Length of variable regions (all samples n=',length(patient.info$Samplename),')',sep=''), subtitle=paste('Wilcox p-value', signif(lenWT$p.value, digits=3) ))

```{r, echo=F, warning=F, message=F, fig.height=5, fig.width=5}
df = rbind(cbind.data.frame('n'=sapply(var.seg.lengths[prog.pt], nrow), 'status'='Progressor'), cbind.data.frame('n'=sapply(var.seg.lengths[np.pt], nrow), 'status'='Non-Progressor'))
nWT = t.test(subset(df, status == 'Progressor')$n, subset(df, status == 'Non-Progressor')$n)

ggplot(df, aes(status, n, group=status)) + geom_jitter(aes(color=status), width=0.2) +
  geom_violin(aes(fill=status, group=status), alpha=0.5) +
  geom_label(data=(df %>% group_by(status) %>% dplyr::summarise(y=max(n), med=median(n))), 
             aes(status, y, label=paste('median\n',sprintf("%.1f", med, sep='')))) +
  labs(y='Number of segments', title="Number of segments per patient", subtitle=paste('t-test p-value', signif(nWT$p.value, digits=3)) ) + theme(legend.position = 'none')
```

## Excluding HGD/IMC

As a consequence of the multipcf segmentation method that jointly segments across all patient samples excluding HGD/IMC samples at this point in the analysis makes no difference in the number/length of segments considered "variable".

```{r echo=F,fig.height=5, fig.width=5}
`%nin%` = Negate('%in%')
nohgd.var.regions = lapply(patient.data, function(pt) {
  #print(unique(pt$info$Hospital.Research.ID))
  nohgd = subset(pt$info, Pathology %nin% c('HGD','IMC'))$Samplename
  if (length(nohgd) > 1) {
    segvals = pt$seg.vals[,c(1:5,grep(paste(nohgd, collapse='|'), colnames(pt$seg.vals)))]

    sds = apply(segvals[,-c(1:5)],1,sd)

    variable.region.indices = which(sds >= sd.threshold & segvals$n.probes >= min.probes)
    variable.regions = segvals[ variable.region.indices,  ]  
  } else {
    variable.regions = pt$variable.regions
  }
  return(variable.regions)
})

nohgd.var.seg.lengths = lapply(nohgd.var.regions, function(vr) {
  cbind('chr'=vr$chrom, 'start'=vr$start.pos, 'end'=vr$end.pos, 'seg.len'=vr$end.pos - vr$start.pos)
})

@@

m = (melt(df, id.vars = 'status', measure.vars = 'seg.len'))
m2 = m %>% group_by(status) %>% dplyr::summarise( median=median(value), n=length(status), value=max(value) )
ggplot(m, aes(status, value, group=status, fill=status)) + geom_jitter(aes(color=status), width=.2) + geom_violin(alpha=0.5) +
  geom_label(data=m2, aes(status, value, label=paste("median=",scientific(median,3),'\nn=',n, sep='')), fill='white') + 
  theme(legend.position = 'none') + labs(y='Variable segment length', x='', title='Length of variable regions excluding HGD/IMC', subtitle=paste('Wilcox p-value', signif(lenWT$p.value, digits=3) ))
```

```{r echo=F,fig.height=5, fig.width=5}
df = rbind(cbind.data.frame('n'=unlist(sapply(nohgd.var.regions[prog.pt], nrow)), 'status'='Progressor'), cbind.data.frame('n'=unlist(sapply(nohgd.var.regions[np.pt], nrow)), 'status'='Non-Progressor'))
nWT = t.test(subset(df, status == 'Progressor')$n, subset(df, status == 'Non-Progressor')$n)

ggplot(df, aes(status, n, group=status)) + geom_jitter(aes(color=status), width=0.2) +
  geom_violin(aes(fill=status, group=status), alpha=0.5) +
  geom_label(data=(df %>% group_by(status) %>% dplyr::summarise(y=max(n), med=median(n))), 
             aes(status, y, label=paste('median\n',sprintf("%.1f", med, sep='')))) +
  labs(y='Number of segments', title="Number of segments per patient, excluding HGD/IMC", subtitle=paste('t-test p-value', signif(nWT$p.value, digits=3)) ) + theme(legend.position = 'none')

@@

```{r echo=F, message=F, warning=F, fig.height=8, fig.width=20}

n.var.regions = sapply(var.seg.lengths, nrow)

@@

#ggplot(melted, aes(Status, value, fill=Status)) + geom_jitter(aes(color=Status), width=0.2) + geom_violin() + 
#  geom_label(data=subset(melted, Status == 'P'), aes(x=Status, y=max(value)-sd(value), label=scientific(median(value),3) ), color='white' , show.legend = F) +
#  geom_label(data=subset(melted, Status == 'NP'), aes(x=Status, y=max(value)-sd(value), label=scientific(median(value),3) ), color='white' , show.legend = F) +
#  labs(title="Variable region length adjusted for chromosome length",subtitle=paste("p-value", signif(chr.ratio.tt$p.value, digits=3)), y='Segment Length/Chr length')
